---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# forwards

[![Travis-CI Build Status](https://travis-ci.org/forwards/forwards.svg?branch=master)](https://travis-ci.org/forwards/forwards)

The *forwards* package provides anonymised data from surveys conducted by [Forwards](http://forwards.github.io/), the R Foundation taskforce on women and other under-represented groups.

## Installation

You can install forwards from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("forwards/forwards")
```

## Usage

The package currently contains a single dataset, `useR2016`, with anonymised results from a survey of participants at the [useR! 2016](http://user2016.org/) conference. The questions and form of responses are described in the help file (`?useR2016`).

The dataset is lazy loaded, so after loading the *forwards* package, the `useR2016` data frame is available for use. For example the following code cross-tabulates respondents' age and length of time using R:

```{r example}
library(forwards)
xtabs(~ Q3 + Q11, data = useR2016)
```

## Code of Conduct

Please note that this package is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.
